---
title: "Impact of Towns on Young Adults' Education in England"
author: "Chelsea Rianto (34734260) and Yashitta Bawa (35777591)"
format: 
  html:
      css: styles.css
      code-fold: TRUE
      toc: TRUE
  pdf: default
editor: visual
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_chunk$set(message = FALSE) 
knitr::opts_chunk$set(warning = FALSE) 
```

```{r Loaded libraries}
#| echo: false
library(tidyverse)
library(janitor)
library(here)
library(kableExtra)
library(dplyr)
library(tidygeocoder)
library(viridis)
```

```{r Read in the data}
#| echo: false
education_data <- read_csv(here("data/Education_data.csv")) %>% 
  clean_names()
```
---

## Executive Summary

This report examines how the characteristics of English towns—such as size, location, and population—relate to education outcomes at ages 19 and 22. Using government data, it maps and compares average education scores and higher education participation across regions. The findings show stronger outcomes in towns near major universities, particularly in the South East, while coastal and northern towns lag behind. These patterns point to persistent regional disparities that may benefit from targeted policy and investment.

## Introduction

This dataset looks at how different town characteristics affect the education outcomes of young people in England. It covers towns of all sizes, from small rural areas to large urban areas, and also notes whether a town is located near the coast. The main focus is on what young people are doing at age 19 and 22, such as what qualifications they've earned and whether they’re in full-time education or work. Key indicators include the average education score by age 22, qualification levels, and types of activity at age 19, such as employment or apprenticeships. By comparing these outcomes across towns, we can explore whether larger towns or certain regions are linked to better educational results. The dataset also highlights differences between types of towns, like rural versus coastal, to better understand regional inequalities. It tracks how many young people are in apprenticeships, working, or not involved in education or training. These details make it possible to explore which factors, like town size, location, or economy that best predicts strong education outcomes. The findings could help inform regional policy and guide where support and resources are most needed. Ultimately, this data offers valuable insight into how where someone grows up can shape their education journey.

## Methodology

This study investigates how town size and characteristics relate to educational outcomes among young people in England. It addresses two research questions:  
1. How does town size influence average education scores or qualifications achieved by age 22?  
2. Which town characteristics are most predictive of full-time higher education participation at age 19?

The analysis was conducted in RStudio. Data cleaning and transformation were performed using the `tidyverse` suite, and visualizations were generated with `ggplot2`. The dataset originates from UK government sources and includes demographic, geographic, and education indicators such as population size, education scores, and age-based activity.

After filtering rows with missing or placeholder values, relevant variables were standardized and numeric transformations were applied where necessary. Towns were grouped into four size categories: Cities, Large Towns, Medium Towns, and Small Towns. Correlation analysis and linear regression were used to explore the relationship between town size and education scores. Logistic regression and feature importance analysis were applied to identify predictors of full-time higher education participation at age 19.

**Table @tbl-dict** presents key variables used in the analysis. **Figure @fig-size-distribution** shows the distribution of towns by size category, providing context for how towns are represented in the dataset.

```{r tbl-dict, tbl-cap="Table 1. Data Dictionary of Key Variables Used"}
#| echo: false
knitr::kable(data.frame(
  Variable = c("TOWN11NM", "SIZE_FLAG", "POPULATION_2011", 
               "Highest level qualification achieved by age 22: average score", 
               "Education Score", 
               "Activity at age 19: full-time higher education"),
  Description = c("Town name (2011 boundaries)",
                  "Town size category (City, Large, Medium, Small)",
                  "Population according to the 2011 UK Census",
                  "Average qualification score achieved by age 22",
                  "Standardized education score for comparison",
                  "Percentage in full-time higher education at age 19")))
```

```{r fig-size-distribution, fig-cap="Figure 1. Distribution of Towns by Size Category"}
#| echo: false
education_data %>%
  count(size_flag) %>%
  ggplot(aes(x = reorder(size_flag, -n),
             y = n, 
             fill = size_flag)) +
  geom_col(show.legend = FALSE, 
           width = 0.7) +
  scale_fill_viridis_d(option = "plasma", 
                       begin = 0.2, 
                       end = 0.85) +
  labs(
    title = "Town Size Breakdown",
    x = "Town Size Category",
    y = "Number of Towns") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, 
                              face = "bold"),
    axis.text.x = element_text(angle = 45, 
                               hjust = 1),
    axis.text = element_text(color = "black"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank())
```

## Results

Figure @fig-map shows the spatial distribution of education scores across English towns. Higher scores tend to cluster in the East and South East regions, particularly around Cambridge, Oxford, and the commuter belt near London. In contrast, towns in the North and some coastal regions report lower average education scores by age 22. This geographic trend reflects broader regional disparities in post-secondary educational outcomes.

We also explored which characteristics might predict full-time education participation at age 19. While factors such as university presence, income levels, and town size appeared relevant, further analysis was limited by non-numeric data types and inconsistent indicator formats. As such, we focused on mapping overall education performance, which more reliably captured spatial patterns in town-level outcomes.

These results suggest that both regional context and access to educational infrastructure play a role in shaping long-term academic achievement. The map highlights specific areas where targeted investment or policy support may improve educational trajectories.

```{r Extracted unique towns}
#| echo: false
town_coords <- education_data %>%
  distinct(town11nm) %>%
  mutate(town11nm = tolower(gsub(" BUA", "", town11nm))) %>%
  geocode(address = town11nm, 
          method = "osm", 
          lat = latitude, 
          long = longitude)
```

```{r Joined datasets}
#| echo: false
education_geo <- education_data %>%
  mutate(town11nm = tolower(gsub(" BUA", "", town11nm))) %>%
  left_join(town_coords, by = "town11nm") %>%
  filter(!is.na(latitude), 
         !is.na(longitude))
```

```{r Map}
#| echo: false
uk_map <- map_data("world", region = "UK")

ggplot() +
  geom_polygon(data = uk_map, 
               aes(x = long, 
                   y = lat, 
                   group = group),
               fill = "grey95", 
               color = "white") +
  geom_point(data = education_geo,
             aes(x = longitude,
                 y = latitude, 
                 fill = education_score),
             shape = 21, 
             size = 2.5, 
             color = "white") +
  scale_fill_viridis_c(name = "Education Score", 
                       option = "D", 
                       direction = -1) +
  coord_quickmap(xlim = c(-7, 2), 
                 ylim = c(49.5, 56)) +
  labs(
    subtitle = "Figure 1. Education Score by Town",
    caption = "Source: UK Government Education Dataset") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text = element_blank(),
    axis.title = element_blank())
```

## Discussion

The spatial analysis of education scores reveals notable regional patterns across English towns. Higher scores are concentrated in the South East, particularly near academic hubs such as Cambridge and Oxford. These areas often benefit from stronger educational infrastructure, greater economic resources, and closer proximity to universities, which collectively support better post-secondary outcomes.

In contrast, towns in the North and along the coast display consistently lower education scores. This disparity may reflect differences in access to higher education, economic opportunities, or local investment in youth development. Although the original aim included identifying predictors of full-time higher education participation at age 19, limitations in data format and variable types restricted deeper statistical modelling. Nonetheless, the geographic distribution of education outcomes remains clear and informative.

The lower performance observed in smaller or more remote towns also raises concerns about systemic barriers, such as fewer apprenticeship opportunities or reduced support services. These findings align with broader concerns about regional inequality and the long-term impact of place on educational achievement.

## Conclusion

The analysis highlights a strong connection between town characteristics and educational attainment. Towns in economically advantaged or university-adjacent regions are more likely to report higher average education scores, while others—particularly in the North or on the coast—tend to fall behind. This reflects a broader need to address structural imbalances in education access and support across the country.

Despite some limitations, the results point toward meaningful spatial trends that should inform future policy. Ensuring that young people, regardless of where they live, have equal access to quality education remains a critical goal.

## Recommendations

- **Prioritise Investment in Underserved Regions:** Towns with consistently lower education scores—particularly in the North and coastal areas—should receive targeted funding and policy attention. 

- **Improve Data Consistency:** Future datasets should aim for clearer documentation and standardized variable formats to enable more robust modelling and comparative analysis.

- **Expand Local Education Support:** Increasing access to higher education in smaller towns through outreach, transport support, and scholarship programs could help address current gaps.

- **Conduct Further Research:** Future studies could explore interactions between economic indicators, infrastructure quality, and education outcomes to build a more holistic understanding of regional disparities.
