---
title: "Assignment 3 Paper"
author: "Chelsea Rianto (34734260) and Yashitta Bawa (35777591)"
format: 
  html:
      toc: TRUE
  pdf: default
editor: visual
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_chunk$set(message = FALSE) 
knitr::opts_chunk$set(warning = FALSE) 
```

```{r Loaded libraries}
library(tidyverse)
library(janitor)
library(here)
library(kableExtra)
```

# Read in Raw Data
```{r Read in the data}
education_data <- read_csv(here("data/Education_data.csv"))
```
---

## Table of Contents

## Executive Summary

## Introduction

This dataset looks at how different town characteristics affect the education outcomes of young people in England. It covers towns of all sizes, from small rural areas to large urban areas, and also notes whether a town is located near the coast. The main focus is on what young people are doing at age 19 and 22, such as what qualifications they've earned and whether they’re in full-time education or work. Key indicators include the average education score by age 22, qualification levels, and types of activity at age 19, such as employment or apprenticeships. By comparing these outcomes across towns, we can explore whether larger towns or certain regions are linked to better educational results. The dataset also highlights differences between types of towns, like rural versus coastal, to better understand regional inequalities. It tracks how many young people are in apprenticeships, working, or not involved in education or training. These details make it possible to explore which factors, like town size, location, or economy that best predicts strong education outcomes. The findings could help inform regional policy and guide where support and resources are most needed. Ultimately, this data offers valuable insight into how where someone grows up can shape their education journey.

## Methodology
This study investigates how the size and characteristics of towns relate to educational outcomes among young people in England. Specifically, it addresses two research questions: 
1. How does town size influence average education scores or qualifications achieved by age 22?
2. Which town characteristics are most predictive of full-time education participation at age 19?

The analysis is conducted in RStudio, data cleaning and transformation are performed with `tidyverse`, and summary statistics and visualizations are produced using `ggpplot2` and `dplyr`. The dataset is from the UK government’s records on towns, demographics, and education indicators, including population size, qualification levels, and employment activity by age group.

After removing rows with missing or placeholder values, numeric transformations are applied where appropriate. Towns are grouped into four size categories: Cities, Large Towns, Medium Towns, and Small Towns. To examine relationships, we will use correlation analysis and linear modeling for the first research question. For the second research question, logistic regression and feature importance ranking will be applied to identify key predictors of full-time education participation at age 19.

Table @tbl-dict outlines the key variables used in the analysis.
```{r tbl-dict, tbl-cap="Table 1. Data Dictionary of Key Variables Used"}
#| echo: false
knitr::kable(data.frame(
  Variable = c("TOWN11NM", "SIZE_FLAG", "POPULATION_2011", 
               "Highest level qualification achieved by age 22: average score", 
               "Education Score", 
               "Activity at age 19: full-time higher education"),
  Description = c("Town name (2011 boundaries)",
                  "Town size category (City, Large, Medium, Small)",
                  "Population according to the 2011 UK Census",
                  "Average qualification score achieved by age 22",
                  "Standardized education score for comparison",
                  "Percentage in full-time higher education at age 19")
))
```

